### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

\newcommand{\gM}{\mathcal{M}}
\newcommand{\gL}{\mathcal{L}}

@@

One of the main ideas we want to investigate in this paper is
how we can use \emph{formulas} from description logics (see~\cite{baad:desc03}) to represent
referring expressions (RE). And how we can use the notion of \emph{bisimulation}
(see~\cite{blac:moda01}) to obtain very efficient algorithms for computing referring expressions.

In what follows we will introduce in detail two description logics called \el and \alc.  Intuitively, \el allows
for conjunctions of atomic and existentially quantified relational expressions (e.g.,
we can write formulas in \el that could be realized as ``a red book on a table''). \alc is the extension of \el with full negation
(e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').
%Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).  On the other hand,
%\el expressions are easily realizable, but the bisimulation algorithm is more complex.

Let start by formally defining the description languages we are going to use, and the notion of bisimulation.

\begin{definition}
Formulas of $\alc$ are generated by the following grammar:
$$
\form ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi' \mid \exists R. \varphi
$$
where $p$ is in the set of propositional symbols \prop, $R$ is in the set of relational symbols \rel, $\varphi$ and $\varphi'$ are in \form. $\el$ is the
negation-free fragment of $\alc$.

Formulas of both $\alc$ and $\el$ are interpreted in relational first-order models
$\gM = (\Delta^\gM,|\cdot|^\gM)$ where $W$ is an arbitrary non-empy set, and $|\cdot|^\gM$ is an interpretation
function such that:
$$
\begin{array}{ccl}
|p|^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
|R|^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
|\neg \varphi|^\gM & = & W \backslash |\varphi|^\gM\\
|\varphi \sqcap \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i' ((i,i') \in |R|^\gM)\\
& & \mbox{ and } i' \in |\varphi|^\gM \}.\\
\end{array}
$$
Given a model $\gM$ and an individual $i$ in the domain of $\gM$, let
$\propm(i) = \{p \in \prop \mid i \in |p|^\gM\}$.
\end{definition}

The previous definition formally introduces the syntax and semantics of the
logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{sec:related}).  For completeness we mention them briefly here,
and we refer the reader to~\cite{baad:desc03} for further details:

\begin{center}
\begin{tabular}{lp{5.9cm}}
$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification.\\
$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification.\\
$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation.
\end{tabular}
\end{center}

\noindent
It should be clear from the definitions above that $\mathcal{CL}$ is conjunctive logic, while $\mathcal{PL}$ is propositional logic.

We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
the case for the $\alc$ language,further details can be found
in~\cite{blac:moda01,kurt:expr99}.

\begin{definition}
Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if there is a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.
\end{itemize}
\end{definition}

Why are bisimulations relevant for computing referring expressions?
The following result provide a crucial characterization~\cite{blac:moda01}:

@@

In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ language. Hence, bisimulation gives us a test for non-referenceability: if a
model $\gM$ contains two different individuals $i$ and $i'$ which are
$\alc$-bisimilar, then there is no formula of $\alc$ that can be used to uniquely
refer to $i$ or $i'$.  We will discuss which is the exact relation between
bisimulation and generation of referring expressions in the next section.
For the moment we just remark that a suitable notion of $\el$-bisimulation can also be defined~\cite{kurt:expr99}.

>>>>

@@

The main point of this paper is to represent referring expressions
(REs) as formulas of description logic \cite{baad:desc03}.  In order
to make this point, we will now first define the two description
logics we will be working with -- \alc\ and \el.

\begin{definition}
Formulas $\varphi$ of $\alc$ are generated by the following grammar:
$$
\varphi,\varphi' ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi'
\mid \exists R. \varphi
$$
where $p$ is in the set of propositional symbols \prop, $R$ is in the
set of relational symbols \rel.

$\el$ is the negation-free fragment of $\alc$.

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta^\gM,|\cdot|^\gM)$ where
$\Delta^\gM$ is some non-empy set, and $|\cdot|^\gM$ is an
interpretation function such that:
$$
\begin{array}{ccl}
|p|^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
|R|^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
|\neg \varphi|^\gM & = & \Delta^\gM \backslash |\varphi|^\gM\\
|\varphi \sqcap \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in |R|^\gM\\
& & \mbox{ and } i' \in |\varphi|^\gM \}.\\
\end{array}
$$
Given a model $\gM$ and an individual $i$ in the domain of $\gM$, let
$\propm(i) = \{p \in \prop \mid i \in |p|^\gM\}$.
\end{definition}

Every formula of a description logic denotes a set of individuals in
the domain; thus we can use such formulas to describe sets.  For
instance, in the model in Fig.~\ref{fig:dale-haddock}b, the formula
$\mathsf{flower}$ denotes the set $\{f_1,f_2\}$; the formula
$\mathsf{flower} \sqcap \exists \mathsf{in}.\mathsf{hat}$ denotes
$\{f_2\}$; and the formula $\mathsf{flower} \sqcap \neg
\exists.\mathsf{hat}$ denotes $\{f_1\}$.

Different description logics differ in the inventory of logical
connectives they permit.  The difference between \alc\ and \el\ is
that \alc\ permits negation and \el\ doesn't.  There are many other
description logics in the literature; some that we will get back to in
Section~\ref{sec:related} are $\mathcal{CL}$ (\el\ without existential
quantification, i.e.\ only conjunctions of atoms); $\mathcal{PL}$
(\alc\ without existential quantification, i.e.\ propositional logic);
and $\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic
negation).

A key notion in studying description logics is \emph{bisimulation}.
We will discuss in detail the case for $\alc$; further details can be
found in~\cite{blac:moda01,kurt:expr99}.

\begin{definition}
Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and
$i'$ of $\Delta^\gM$ are \emph{$\alc$-bisimilar} if there is a relation $Z
\subseteq \Delta^\gM \times \Delta^\gM$ such that 
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
$(e_1',e_2') \in Z$.
\end{itemize}
A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
\end{definition}

The crucial property that makes bisimulation relevant for GRE is the
following theorem \cite{blac:moda01}.

@@

In other words, $\alc$-bisimilar individuals are indistinguishable in
the $\alc$ language.  In the example in Fig.~\ref{fig:dale-haddock}a,
the two individuals $t_1$ and $t_2$ are \alc-bisimilar; and indeed,
they satisfy exactly the same \alc-formulas.  Although we can't go
into details here, it is possible to define a suitable notion of
$\el$-bisimulation for which the analogous theorem holds
\cite{kurt:expr99}.  In general, \el-bisimulation is a weaker notion
than \alc-bisimulation; for instance, in Fig.~\ref{fig:dale-haddock},
$f_1$ and $f_2$ are \el-bisimilar \todo{oops! here's the asymmetry,
  there is a formula for $f_2$ but not for $f_1$} (no \el-formula will
distinguish them), but we have already seen that they satisfy
different \alc-formulas.

### 2008-03-20 trunk/Papers/dl-gre-08/discussion.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

\subsection{Interface to realization}

@@

Our GRE algorithms do not guarantee that the concepts they compute can
actually be realized in language.  For example, none of the concepts
our algorithms computed in the Viethen \& Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing the concept $\neg
\exists R.(\neg P \sqcap \neg Q)$ from the example in
Fig.~\ref{fig:el-vs-alc} as a smooth noun phrase.  Although we agree
with \newcite{deemter01:_gener_refer_expres} and others that the
careful use of negation and disjunction can improve REs, these
connectives must not be overused.  Thus we consider the concepts
computed by the \el\ algorithm ``safer'' with respect to realization.

>>>>

\subsection{Interface to realization}
\label{sec:discussion-realization}

@@

Our GRE algorithms do not guarantee that the concepts they compute can
actually be realized in language.  For example, none of the concepts
our algorithms computed in the Viethen \& Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing a concept like
$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
Although we agree with \newcite{deemter01:_gener_refer_expres} and
others that the careful use of negation and disjunction can improve
REs, these connectives must not be overused.  Thus we consider the
concepts computed by the \el\ algorithm ``safer'' with respect to
realization.

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

In computational linguistics, the GRE
problem can be intuitively presented as follows: given some information about
a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
and a grammar, find a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.

By taking a logic perspective, we can recast the GRE problem as an inference problem.
We can assume that the information about the group of individuals is provided in the form
of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
associated with the GRE problem will be to find a formula in $\gL$ such that it
exaclty identifies a given individual $i$.
\medskip

\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and an individual $i$ in the\\
& \hspace*{0.5cm} domain of $\gM$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

As it can be seen from the definition, the problem depends on which particular
logical language $\gL$ we are interested in, and (as the original GRE problem)
it might terminate with failure.  Depending on \emph{a)} the model $\gM$,
\emph{b)} the chosen individual $i$
and \emph{c} the expressive power of $\gL$, it can be the case that there is no formula of
$\gL$ that uniquely identifies $i$ in $\gM$.

Consider the following example.  We have three individuals related as indicated in
Fig.~\ref{fig:dale-haddock}d.

%\begin{figure}
%\begin{center}
%\includegraphics[scale=.8]{pic.pdf}
%\end{center}
%\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
%\end{figure}
%
The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
referring expression for $i$ would be ``an individual relating to itself''.
But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
are ``individuals which are related to individuals which are related to individuals\ldots''

We can connect description logics to the problem of generating
referring expressions via the concept of \emph{bisimulation classes}.
A bisimulation class is a maximal subset $C$ of the domain such that
all members of $C$ are pairwise bisimilar.  For the finite models we
deal with in GRE, this amounts to saying that all individuals in $C$ and only those satisfy the same formulas (Theorems~\ref{bisim}).  If we represent referring expressions as description logic formulas,
this means that there exists a unique referring expression for some
individual $i$ iff $i$ is alone in its bisimulation class; and if it
is, we can use one of the formulas that are characteristic for $C$ as
the referring expression.

Representing referring expressions as DL concepts is a very natural
perspective.  The purely conjunctive set of propositional symbols that,
e.g.,~\newcite{Dale1995} compute is clearly a conjunction in
description logic.  Relational expressions as in
\newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
We will discuss this idea further in Section~\ref{sec:related}.

Summing up, we can reduce the problem of computing a unique RE
to the problem of computing the bisimulation classes of a model.  We
will now present two algorithms -- one for \alc\ and one for \el\ --
for doing this.  In fact, each of these algorithms computes all
bisimulation classes of a model at once, effectively computing a RE
for every single individual in the domain simultaneously.

@@

There exists very efficient algorithms
for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
These algorithms were originally used for automata minimization (given an
automata $A$ they would return a smaller automata $A'$ which recognized the
same language), but we can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
discuss the $\alc$ case.

We assume given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$ containing all the information about the current
situation.  We will compute a set $\RE$ of $\alc$ formulas such that the
set $\{|\varphi|^\gM \mid \varphi \in \RE\}$ is a partition of $\Delta^\gM$
that contains all the bisimulation classes of $\gM$.  The pseudocode is
shown in Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.

\begin{theorem}\label{theo:bisimulation-classes}
Given a model $\gM$, the set $\mathcal{C}$ of all $\alc$-bisimulation classes
of $\gM$ is computed by Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-alc} in polynomial time.  Moreover, for each class $C \in \mathcal{C}$ the algorithm returns an $\alc$ formula $\varphi$ such that
$C = |\varphi|^\gM$.
\end{theorem}

@@

Algorithms~\ref{algo:bisim-l} and \ref{algo:bisim-add-alc} compute
the \alc-bi\-sim\-u\-la\-tion classes for a given model.
Algorithm~\ref{algo:bisim-l} iterates over all propositional and
relational symbols of the signature to construct new formulas until
either all formulas in $\RE$ denote singletons (i.e., there is only one
individual that satisfies them), or no progress has been made in the
previous iteration.  In each iteration, it calls the procedure
add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any formula $\psi
\in \RE$ which does not denote a singleton and which is
not equivalent to $\varphi$ and to $\neg \varphi$. In this case, it
replaces $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
$\RE$.

The $\alc$ algorithm is driven by case distinctions between
concepts $\varphi$ and $\neg \varphi$, and thus freely adds 
negations to the characteristic formulas in $\RE$.  This can be inconvenient
in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{sec:examples}.

An algorithm which is limited to computing
(negation-free) formulas in \el can be obtained by combining
Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-el}.

As before, the algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
formulas (this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines iteratively.  However,
where the \alc\ algorithm maintains the invariant that
$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of $\Delta$, we
weaken this invariant to the requirement that there are no $m \geq 2$
pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
$\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup \ldots \cup
\interp{\varphi_{i_m}}$.  We call the formula $\varphi_{i_1}$ \emph{subsumed} if
such a decomposition exists.

The procedure add$_\el$($\varphi$, $\RE$) shown in Algorithms~\ref{algo:bisim-add-el} intersects $\varphi$ with all current formulas in $\RE$
that do not denote a singleton, adds the result to $\RE$ and then
removes all formulas in $\RE$ that have become subsumed.  Removal of
subsumed formulas from $\RE$ can be implemented efficiently by arranging
them by inclusion in terms of their denotation.

Because the \el\ algorithm maintains a weaker invariant
for the classes, it may consider more of these classes simultaneously.
Prima facie, this means that it has worst-case exponential runtime,
given that the whole domain has an exponential number of subsets.
However, it is possible that a more careful analysis of the
combinatorics of the considered subsets reveal that the algorithm is
actually polynomial; it certainly behaves polynomially in our
evaluations (see below).

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{pic-dale-haddock}
  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
    scenario; (b) the \newcite{Stone1998a} scenario; (c) illustrating
    the difference between \el\ and \alc; (d)
    distinguishable/indistinguishable points.}
  \label{fig:dale-haddock}
\end{figure*}

@@

In the first loop,
the formulas $\mathsf{floor}$, $\mathsf{bowl}$,
$\mathsf{cup}$, and $\mathsf{table}$ will be added to $\RE$ because they subsume $\top$.  Not all of these formulas denote singletons; for instance,
$|\mathsf{cup}|\gM$ contains two individuals.  So we iterate over
the relations to refine our formulas.  After the first
iteration over the relations, we have $\RE = \{ \mathsf{floor},
\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup},
\mathsf{table} \}$. Notice that $\mathsf{bowl}$ has become subsumed,
but we haven't distinguished the cups and tables further.

@@

At this point, all
formulas except $\mathsf{table}$ denote singletons, and further
iterations don't allow us to refine $\mathsf{table}$; so the algorithm
terminates.

Each of the singleton bisimulation classes is represented
by a characteristic formula; for instance, $\RE$ tells us that
$\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap
\exists \mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we
may refer to $c_2$ as ``the cup in the bowl on the table''.

Compared to the \alc\ algorithm, the \el\ algorithm will generally
compute larger bisimulation classes, and compute singleton
bisimulation classes less frequently, because the concepts it can
build can ``see'' fewer differences between individuals.  For example,
in the model shown in Fig.~\ref{fig:dale-haddock}c, the \alc algorithm
will compute the set $\RE = \{\neg \exists R. (\neg P \sqcap \neg Q),
Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each formula denotes a singleton), but $a$ and $c$ are \el-bisimilar in this model,
and the only \el-concept that is true in $d$ is $\top$, so the \el\
algorithm can only compute the formulas $\{\top, P, P \sqcap Q\}$.

>>>>

Now we apply description logic to GRE.  The core claim of this paper
is that the GRE problem can be viewed as the problem of computing a
formula of description logic whose extension is exactly a given set of
individuals.

\medskip
\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = A$\\
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

In the examples above, it is because $\mathsf{flower} \sqcap \exists
\mathsf{in}. \mathsf{hat}$ denotes exactly $\{f_2\}$ that we can say
``the flower in the hat'' to refer to $f_2$.  This generalizes the
common view of GRE as attribute selection \cite{Dale1995}: From our
perspective, a bag of attributes is simply a conjunction of atoms.
However, more expressive approaches to GRE fit into our framework as
well; for example, relational expressions as in
\newcite{dale91:_gener_refer_expres_invol_relat} are formulas of \el.
Different approaches will deliver different REs, and indeed will be
able to build REs at all in different circumstances; this reflects the
differences between the DL languages they use to represent their REs
in.  This idea provides a unified perspective for comparing different
GRE approaches, and we will discuss how to see several further
approaches in Section~\ref{sec:related}.

For the rest of this paper, we assume that we are generating a
singular RE, i.e.\ $A$ will generally be a singleton set.  In this
case, we will only be able to generate a formula that denotes exactly
$A = \{a\}$ (i.e., a RE that uniquely refers to $a$) iff there is no
other individual $b$ that is bisimilar to $a$; otherwise, any formula
that is satisfied by $a$ is also satisfied by $b$.  This is why there
is no RE that uniquely refers to $t_2$ in Fig.~\ref{fig:dale-haddock}a
and only uses the relations ``on'' and ``in'' (and not their
inverses).  In other words, $A$ must be a bisimulation class.
Conversely, if we know that $a$ is alone in its bisimulation class, we
know that there is a concept that describes only $a$ and nothing
else.  This means that we can recast the problem of generating
referring expressions as the problem of computing the bisimulation
classes of a model.

In the rest of this section, we will present algorithms that compute
the bisimulation classes of a given model for \alc\ and \el, along
with a characteristic formula for each class.  These algorithms
efficiently compute \emph{all} bisimulation classes of a model, and
not just the one for some target referent.  In effect, they compute
REs for all individuals in some model at the same time.  The REs
generated by the \el\ algorithm correspond to the relational REs
generated e.g.\ by \newcite{dale91:_gener_refer_expres_invol_relat}
(but without any danger of infinite regress), whereas the REs
generated by the \alc\ algorithm can freely use negation as well.

@@

There is substantial literature on the topic of computing the
\alc-bisimulation classes of a given model efficiently
\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
However, these algorithms will only compute the bisimulation classes
themselves. We extend the \newcite{hopc:algo71} algorithm in such a
way that it computes, for each bisimulation class, also a formula that
denotes exactly this class.  We can then use these formulas as
representations of the REs.

The pseudocode for our \alc\ algorithm is shown as
Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
Given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$, the algorithm computes
a set $\RE$ of $\alc$ formulas such that the set $\{|\varphi|^\gM \mid
\varphi \in \RE\}$ is the set of \alc-bisimulation classes of $\gM$.
The algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
\Delta^\gM$), and successively refines $\RE$ by making its elements
denote smaller and smaller sets.  It maintains the invariant that at
the start and end of every iteration, $\{|\varphi|^\gM \mid \varphi
\in \RE\}$ is always a partition of $\Delta^\gM$.  The algorithm
iterates over all propositional and relational symbols of the
signature to construct new formulas until either all formulas in $\RE$
denote singletons (i.e., there is only one individual that satisfies
them), or no progress has been made in the previous iteration.  In
each iteration, it calls the procedure add$_\alc$($\varphi$, $\RE$),
which intersects $\varphi$ with any formula $\psi \in \RE$ which does
not denote a singleton and which is not equivalent to $\varphi$ and to
$\neg \varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi
\sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to $\RE$.

@@

The \alc\ algorithm is capable of computing the \alc-bisimulation
class of the model very efficiently, in time $O(nm)$ where $n$ is the
number of individuals in the domain, and $m$ is the number of pairs of
individuals that are related by some relation.  However, it will
freely introduce negations in the case distinctions, which can make
the resulting formula hard to realize (see also
Section~\ref{sec:discussion-realization}).  This is why we also
present an algorithm for the \el-bisimulation classes; \el\
corresponds to positive relational REs, which are generally easy to
realize.

We obtain the \el\ algorithm by replacing the call to add$_{\alc}$ in
Algorithm~\ref{algo:bisim-l} by a call to add$_{\el}$, which is defined
in Algorithm~\ref{algo:bisim-add-el}.  As before, the algorithm
maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of formulas
(this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines
iteratively.  However, where the \alc\ algorithm maintains the
invariant that $\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a
partition of $\Delta$, we weaken this invariant to the requirement
that there are no $m \geq 2$ pairwise different indices $1 \leq
i_1,\ldots,i_m \leq n$ such that $\interp{\varphi_{i_1}} =
\interp{\varphi_{i_2}} \cup \ldots \cup \interp{\varphi_{i_m}}$.  We
call the formula $\varphi_{i_1}$ \emph{subsumed} if such a
decomposition exists.

Because the \el\ algorithm maintains a weaker invariant for the
classes, it may consider more of these classes simultaneously.  Prima
facie, this means that it has worst-case exponential runtime, given
that the whole domain has an exponential number of subsets.  However,
it is possible that a more careful analysis of the combinatorics of
the considered subsets will reveal that the algorithm is actually
polynomial.

@@

The algorithm starts with $\RE = \{\top\}$.  In the first loop, it
adds the formulas $\mathsf{floor}$, $\mathsf{bowl}$, $\mathsf{cup}$,
and $\mathsf{table}$, and then removes $\top$ because it is now
subsumed.  Not all of these formulas denote singletons; for instance,
$|\mathsf{cup}|^\gM$ contains two individuals.  So we iterate over the
relations to refine our formulas.  After the first iteration over the
relations, we have $\RE = \{ \mathsf{floor}, \mathsf{bowl} \sqcap
\exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
\mathsf{on}.\mathsf{table}, \mathsf{cup}, \mathsf{table} \}$. Notice
that $\mathsf{bowl}$ has become subsumed, but we haven't distinguished
the cups and tables further.

@@

At this point, all formulas except $\mathsf{table}$ denote singletons,
and further iterations don't allow us to refine $\mathsf{table}$; so
the algorithm terminates.  Each of the singleton bisimulation classes
is represented by a characteristic formula; for instance, $\RE$ tells
us that $\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl}
\sqcap \exists \mathsf{on}.\mathsf{table})$ is only satisfied by
$c_2$, so we may refer to $c_2$ as ``the cup in the bowl on the
table''.  Notice that the algorithm didn't focus on any particular
individual; it simultaneously generated REs for all individuals except
for the two tables.

\begin{figure}[t]
  \centering
  \includegraphics[width=\columnwidth]{pic-dale-haddock}
  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
    scenario; (b) the \newcite{Stone1998a} scenario.}
  \label{fig:dale-haddock}
\end{figure}

The \el\ algorithm has a harder time with the example in
Fig.~\ref{fig:dale-haddock}b \cite{Stone1998a}.  While it will
correctly identify $r_1$ as ``the rabbit in the hat'' and $f_2$ as
``the flower in the bathtub'', it will not be able to compute a
RE for $f_1$; indeed, the algorithm terminates with $\RE$ containing
both $\mathsf{flower}$ and $\mathsf{flower} \sqcap \exists
\mathsf{in}.\mathsf{hat}$.  By contrast, the \alc\ algorithm will
split $\mathsf{flower}$ into the two new formulas $\mathsf{flower}
\sqcap \exists \mathsf{in}.\mathsf{hat}$ and $\mathsf{flower} \sqcap
\neg \exists \mathsf{in}.\mathsf{hat}$, generating a unique RE for
$f_1$ as well.

### 2008-03-20 trunk/Papers/dl-gre-08/introduction.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

The generation of referring expressions is one of the most active and
successful research areas in natural language generation.  Starting with Dale and Reiter's
incremental algorithm \cite{Dale1995}, various researchers have added
extensions such as reference to sets \cite{Stone2000}, more expressive
logical connectives \cite{deemter01:_gener_refer_expres}, and
relational expressions \cite{dale91:_gener_refer_expres_invol_relat}.

>>>>

The generation of referring expressions (GRE) is one of the most
active and successful research areas in natural language generation.
Starting with Dale and Reiter's incremental algorithm \cite{Dale1995},
various researchers have added extensions such as reference to sets
\cite{Stone2000}, more expressive logical connectives
\cite{deemter01:_gener_refer_expres}, and relational expressions
\cite{dale91:_gener_refer_expres_invol_relat}.
